\documentclass[11pt]{article}
\input{docs/manuals/03_src/latex/_preamble.tex}

\title{CX Manual (Master)}
\author{}
\date{Last updated: 2026-02-26}

\begin{document}
\maketitle
\vspace{-0.6em}

\section*{Intent}
cx is built for one outcome: \textbf{predictable LLM-assisted work under load}.
That means bounded context, schema-validated structured outputs, strict telemetry, and safe
automation boundaries.

This master manual synthesizes three viewpoints:
\begin{itemize}
  \item a story-first walkthrough (to build the mental model),
  \item a field guide (to debug and tune quickly),
  \item an operator playbook (to validate and execute safely).
\end{itemize}

\section*{Architecture Overview}
\begin{itemize}
  \item Canonical engine: \texttt{cxrs} (Rust). Bash is fallback only.
  \item Authoritative entrypoint: \texttt{bin/cx} routes Rust-first with explicit fallback.
  \item Capture pipeline applies to \emph{system output} only (optional RTK, then budgets).
  \item Structured commands are schema-enforced (JSON-only, validated, replayable).
  \item Schema failures are quarantined (\texttt{.codex/quarantine/}) and logged with \texttt{quarantine\_id}.
  \item Run telemetry is append-only JSONL (\texttt{.codex/cxlogs/runs.jsonl}).
  \item Policy engine blocks destructive/out-of-repo writes by default.
\end{itemize}

\section*{Technical Expose (Rust Refactor Branch)}
\begin{itemize}
  \item Active branch focus: split monolithic command logic into cohesive Rust modules.
  \item Orchestrator: \texttt{src/app/mod.rs} (routing + composition).
  \item Runtime config is centralized in \texttt{src/modules/config.rs} (\texttt{AppConfig} loaded once at startup).
  \item Core extracted command modules: \texttt{introspect}, \texttt{runtime\_controls}, \texttt{agentcmds}, \texttt{logview}, \texttt{analytics}, \texttt{diagnostics}, \texttt{routing}, \texttt{prompting}, \texttt{optimize}, \texttt{doctor}, \texttt{schema\_ops}, \texttt{settings\_cmds}, \texttt{structured\_cmds}, \texttt{task\_cmds}.
  \item Non-schema LLM commands share one execution path in \texttt{agentcmds} via \texttt{execute\_llm\_command(..., LlmMode)}.
  \item Behavioral contract is unchanged: deterministic schemas, quarantine/replay, policy gating, budgeted capture, append-only logs.
  \item Quality gate per extraction slice: \texttt{cargo fmt} + \texttt{cargo test} must pass.
\end{itemize}

\section*{Quickstart (3 Commands)}
\begin{lstlisting}[style=cx]
cd <repo-root>
./bin/cx version
./bin/cx budget
./bin/cx diffsum-staged | jq .
\end{lstlisting}

\section*{Common Workflows (Copy/Paste)}
\begin{lstlisting}[style=cx]
# Structured outputs:
./bin/cx diffsum-staged | jq .
./bin/cx commitjson | jq .

# Task graph:
./bin/cx task fanout "Objective..." --from staged-diff
./bin/cx task run-all --status pending

# Optimization:
./bin/cx optimize 200
\end{lstlisting}

\section*{Installation / Prereqs}
\begin{itemize}
  \item Required: \texttt{bash}
  \item Required: \texttt{git}
  \item Required: \texttt{jq}
  \item Required: \texttt{codex} CLI (default backend)
  \item Development: Rust toolchain (\texttt{cargo}, \texttt{rustc})
\end{itemize}

\begin{lstlisting}[style=cx]
./bin/cx doctor
./bin/cx rtk-status
\end{lstlisting}
Optional: \texttt{rtk} (system output only), \texttt{ollama} (local backend).

\section*{Troubleshooting (Fast)}
\begin{itemize}
  \item Unexpected fallback: \texttt{./bin/cx where <cmd>} and confirm \texttt{execution\_path} via \texttt{./bin/cx version}.
  \item Schema command fails: inspect \texttt{./bin/cx quarantine list}, then \texttt{./bin/cx replay <id>}.
  \item Budget clipping surprises: check \texttt{./bin/cx budget} and \texttt{./bin/cx trace}; reduce capture scope or adjust budgets intentionally.
  \item Strict log validation fails: legacy rows may predate the current telemetry contract; new runs should conform.
  \item Backend confusion: run \texttt{./bin/cx llm show} and verify the selected model when using ollama.
\end{itemize}

\tableofcontents
\newpage

\section{Start Here: Verify What You’re Running}
\subsection*{Fast path (30 seconds)}
\begin{lstlisting}[style=cx]
cd <repo-root>
./bin/cx version
./bin/cx diag
./bin/cx where version diffsum-staged task optimize
\end{lstlisting}

If \texttt{execution\_path} is not Rust, you are on fallback. Expect reduced guarantees.

\section{The Pipeline (What cx Does With Your Reality)}
\subsection{Capture and Budgeting}
The core reliability move is this: \textbf{system output is constrained before it becomes prompt text}.

\begin{lstlisting}[style=cx]
raw system output
  -> optional RTK routing (system output only)
  -> optional native reduction
  -> mandatory budgeting (clip/chunk)
  -> embed into prompt
\end{lstlisting}

Budgets (defaults):
\begin{itemize}
  \item \texttt{CX\_CONTEXT\_BUDGET\_CHARS=12000}
  \item \texttt{CX\_CONTEXT\_BUDGET\_LINES=300}
  \item \texttt{CX\_CONTEXT\_CLIP\_MODE=smart|head|tail} (default \texttt{smart})
  \item \texttt{CX\_CONTEXT\_CLIP\_FOOTER=1}
\end{itemize}

Inspect current budgets + last-run clip stats:
\begin{lstlisting}[style=cx]
./bin/cx budget
./bin/cx trace
\end{lstlisting}

\subsection{Modes}
\begin{itemize}
  \item \texttt{lean}: minimal prompts/output
  \item \texttt{deterministic}: strict schema-only behavior where applicable
  \item \texttt{verbose}: additional human detail (where supported)
\end{itemize}

Schema commands always force deterministic mode by default. Override with \texttt{CX\_SCHEMA\_RELAXED=1}.

\section{Structured Outputs (Schemas, Quarantine, Replay)}
\subsection{Schema Registry}
Schemas live under \texttt{.codex/schemas/}:
\begin{itemize}
  \item \texttt{commitjson.schema.json}
  \item \texttt{diffsum.schema.json}
  \item \texttt{next.schema.json}
  \item \texttt{fixrun.schema.json}
\end{itemize}

\begin{lstlisting}[style=cx]
./bin/cx schema list
./bin/cx schema list --json | jq .
\end{lstlisting}

\subsection{Schema Commands}
\begin{lstlisting}[style=cx]
./bin/cx commitjson | jq .
./bin/cx diffsum-staged | jq .
./bin/cx next git status | jq .
\end{lstlisting}

\subsection{Failure Procedure}
If validation fails, cx writes the raw response + metadata to quarantine and returns non-zero.

\begin{lstlisting}[style=cx]
./bin/cx quarantine list
./bin/cx quarantine show <id>
./bin/cx replay <id>
\end{lstlisting}

\section{Safety Sandbox (Policy Engine)}
The policy engine is the boundary between ``suggest'' and ``execute''.

Blocked by default (minimum):
\begin{itemize}
  \item \texttt{sudo}, \texttt{rm -rf}, \texttt{curl | bash/sh/zsh}
  \item \texttt{chmod/chown} on system paths
  \item writes outside repo root
\end{itemize}

\begin{lstlisting}[style=cx]
./bin/cx policy show
./bin/cx policy check "rm -rf ."
\end{lstlisting}

\section{Task Graph (Fan-out, Run, Run-all)}
\subsection{Why tasks}
Tasks are the unit of migration to parallel work: small, role-tagged, log-linked, and replayable.

\begin{lstlisting}[style=cx]
./bin/cx task fanout "Ship release notes improvements" --from staged-diff
./bin/cx task list --status pending
\end{lstlisting}

\subsection{Run tasks}
\begin{lstlisting}[style=cx]
./bin/cx task run task_001 --mode deterministic --backend codex
./bin/cx task run-all --status pending
\end{lstlisting}

\section{Telemetry (Contract, Validation, Interpretation)}
\subsection{Log locations}
\begin{itemize}
  \item runs: \texttt{.codex/cxlogs/runs.jsonl}
  \item schema failures: \texttt{.codex/cxlogs/schema\_failures.jsonl}
\end{itemize}

\subsection{Validate}
\begin{lstlisting}[style=cx]
./bin/cx logs validate --fix=false
\end{lstlisting}
Note: strict validation will flag older historical rows that predate the current contract.

\subsection{Interpretation (fast heuristics)}
\begin{itemize}
  \item \texttt{effective\_input\_tokens} high: prompts too large or too variable.
  \item cache trend down: stabilize prompt templates and capture scope.
  \item frequent clipping: reduce capture scope or raise budgets intentionally.
  \item schema failures spike: enforce deterministic mode and reduce context ambiguity.
\end{itemize}

\section{Self-Optimization (Pure Analysis)}
\begin{lstlisting}[style=cx]
./bin/cx optimize 200
./bin/cx optimize 200 --json | jq .
\end{lstlisting}

\section{Operator Procedures (Copy/Paste)}
\subsection*{Budget stress test}
\begin{lstlisting}[style=cx]
CX_CONTEXT_BUDGET_CHARS=2000 CX_CONTEXT_BUDGET_LINES=40 ./bin/cx cxo git status
./bin/cx budget
./bin/cx trace
\end{lstlisting}

\subsection*{Schema integrity check}
\begin{lstlisting}[style=cx]
./bin/cx commitjson | jq .
./bin/cx diffsum-staged | jq .
\end{lstlisting}

\subsection*{Task loop}
\begin{lstlisting}[style=cx]
./bin/cx task fanout "Objective: tighten schema errors" --from log
./bin/cx task run-all --status pending
./bin/cx optimize 200
\end{lstlisting}

\end{document}
